---
title: "Natural variation of flight"
author: "Kyu Min Huh"
output:
  html_document:
    code_folding: show
    fig_height: 5
    fig_width: 7
    toc: TRUE
    toc_depth: 5
    toc_float: 
      collapsed: false
      smooth_scroll: false
editor_options:
  chunk_output_type: console
date: "2025-11-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Summary

We analysed 1434 flights of 49 individuals of blue tit, great tit, and marsh tit to identify interspecific differences in take-off flight performance. We found significant difference in initial velocity and in-flight acceleration between species. Great tits showed higher initial velocity and lower in-flight acceleration than blue tits. Marsh tits did not differ in flight performance with other species. Weather conditions did not show significant effect to flight performance. 

# I. Data 
## libraries 
```{r libraries, warning = FALSE, message = FALSE}
library(dplyr)
library(stringr)
library(readr)
library(fuzzyjoin)
library(tidyverse)
library(ggplot2)
library(lubridate)
library(ggpubr)
library(rstatix)
library(coin)
library(scales)
library(gridExtra)
library(ggthemes)
library(ggbreak)
library(lme4)
library(reshape2)
library(olsrr)
library(gvlma)
library(PerformanceAnalytics)
library(usethis)
library(devtools)
library(redres)
library(EnvStats)
library(lmtest)
library(lmerTest)
library(AICcmodavg)
library(emmeans)
library(MuMIn)
library(sjPlot)
library(rptR)
```

## dataset 
```{r}

all_dates3 <- read.delim("C:\\Users\\jesu4036\\Documents\\wytham-tits-fly\\all_dates3.txt", sep = ",") %>% distinct()

all_dates3$datetime.y <- as.POSIXct(all_dates3$datetime.y, format = "%Y-%m-%d %H:%M:%S")
colnames(all_dates3)
all_dates3 <- all_dates3 %>% 
  select(-Spec)

climate <- read.delim(file = "C:\\Users\\jesu4036\\Documents\\wytham-tits-fly\\climate_filled.csv", sep = ',')

climate$datetime <- as.POSIXct(climate$datetime, format = "%Y-%m-%d %H:%M:%S")

caught.hour.day <- read.delim(file = "C:\\Users\\jesu4036\\Documents\\wytham-tits-fly\\caught.hour.day.csv", sep = ',')

```

## adding day and hour and climate 
```{r}
all_dates3$day <- day(all_dates3$datetime.y)
all_dates3$hour <- hour(all_dates3$datetime.y)

all_dates3$roundate <- round_date(all_dates3$datetime.y, "10 mins")
all_dates3.climate <- merge(x = all_dates3, y = climate, by.x = "roundate", by.y = "datetime", all.x = TRUE)

flight.data <- all_dates3.climate %>%
  filter(day %in% c(15:24))

nat.var <- all_dates3.climate %>%
  filter(day %in% c(15:19))

capt.eff <- all_dates3.climate %>%
  filter(day %in% c(20:24))


```

write.csv(nat.var, "C:\\Users\\kmh\\Desktop\\DATA\\nat.var.csv")

## data synopsis {.tabset}
### raw continuous data 
```{r}

dim(flight.data) #total number of flight 
table(flight.data$day)

no.of.individuals <- rowSums(table(flight.data$species, flight.data$Ring)!=0) 
no.of.flights <- table(flight.data$species)

raw_sum <- rbind(no.of.individuals, no.of.flights)

raw_sum


```

### natural variation data
```{r}
dim(nat.var) #total number of flight 
table(nat.var$day)

no.of.individuals <- rowSums(table(nat.var$species, nat.var$Ring)!=0) 
no.of.flights <- table(nat.var$species)

nat.var_sum <- rbind(no.of.individuals, no.of.flights)

nat.var_sum

breeding <- cbind(nat.var$Ring, nat.var$RFID, nat.var$species, nat.var$age, nat.var$Sex) %>% as.data.frame() %>%
  distinct()
colnames(breeding) <- c("Ring", "RFID", "species", "age", "Sex")

```
write.csv(breeding, "\\breeding.csv",row.names = FALSE )


### For each species, M vs. F, young vs. adult, then between species
```{r}
#find the ring number of those without Sex in Great Tit & Blue Tit 
for( i in 1:nrow(nat.var)){
  
  if(nat.var$Ring[i] == "ATR8387"){
    
    nat.var$Sex[i] <- "F"
  }
  
  else if(nat.var$Ring[i] == "ATR8383"){
    
    nat.var$Sex[i] <- "F"
  }
  else if(nat.var$Ring[i] == "PF72506"){
    
    nat.var$Sex[i] <- "M"
  }
  else{}
  
}

#Sex unknown for ATR8380 and ATR5368 both in morphometrics and brood data
nosex <- nat.var %>%
  filter(species != "marti") %>%
  filter(Sex == "") %>%
  select(Ring) %>%
  distinct()  
```

### no.of individuals by species, age, sex in natural variation
```{r}
numbers <- nat.var %>%
  count(species, age, Sex, Ring)


individual.age <- numbers %>%
  count(species, age)

individual.sex <- numbers %>%
  count(species, Sex)

numbers.age <- numbers %>%
   group_by(species, age) %>%
   summarise(mean = mean(n), sd = sd(n)) %>%
             na.omit()

numbers.sex <- numbers %>%
  group_by(species, Sex) %>%
  summarise(mean = mean(n), sd = sd(n)) %>%
            na.omit()

numbers.age <- numbers.age %>%
  add_column(individual.age$n)

numbers.sex <- numbers.sex %>%
  add_column(individual.sex$n)

numbers.tot <- numbers %>%
  group_by(species) %>%
  summarise(mean = mean(n), sd = sd(n)) %>%
            na.omit()

numbers.age
numbers.sex
numbers.tot
```

#### t-tests on visiting rate
```{r}

bluti <- numbers %>%
  filter(species == "bluti")

greti <- numbers %>%
  filter(species == "greti")

marti <- numbers %>%
  filter(species == "marti")



bluti.F <- numbers %>%
   filter(species == "bluti" & Sex == "F")

bluti.M <- numbers %>%
   filter(species == "bluti" & Sex == "M")

bluti.5 <- numbers %>%
   filter(species == "bluti" & age == 5)

bluti.6 <- numbers %>%
   filter(species == "bluti" & age == 6)

greti.F <- numbers %>%
  filter(species == "greti" & Sex == "F")

greti.M <- numbers %>%
  filter(species == "greti" & Sex == "M")

greti.5 <- numbers %>%
  filter(species == "greti" & age == 5)

greti.6 <- numbers %>%
  filter(species == "greti" & age == 6)

t.test(bluti.F$n, bluti.M$n, paired = FALSE, var.equal = FALSE, conf.level = 0.95)
t.test(bluti.5$n, bluti.6$n, paired = FALSE, var.equal = FALSE, conf.level = 0.95)
t.test(greti.F$n, greti.M$n, paired = FALSE, var.equal = FALSE, conf.level = 0.95)
t.test(greti.5$n, greti.6$n, paired = FALSE, var.equal = FALSE, conf.level = 0.95)

t.test(bluti$n, greti$n, paired = FALSE, var.equal = FALSE, conf.level = 0.95)
t.test(bluti$n, marti$n, paired = FALSE, var.equal = FALSE, conf.level = 0.95)
t.test(greti$n, marti$n, paired = FALSE, var.equal = FALSE, conf.level = 0.95)
```

#### headwind, air temperature, humidity
```{r}

#overall climate data 
#find how many weather values are missing
#missing data that do not make to the final dataset. 
View(flight.data[!complete.cases(flight.data),])
#mean weather 

mean(flight.data$air.temp, na.rm = TRUE)
mean(flight.data$headwind, na.rm = TRUE) #relative to the birds flying out of the tunnel
mean(flight.data$humidity, na.rm = TRUE)


#dataset for baseline variation 
climate.nat.var <- nat.var %>%
  filter(day %in% c(15:19)) %>%
  select(headwind, air.temp, humidity)
  
#not using caught.hour.day which is a reduced dataset to just answer the question if the climate conditions were different at nat.var vs. capt.eff
#otherwise everything is significant probably the sample size is so vastly different
climate.capt.eff <- caught.hour.day %>%
  select(headwind, air.temp, humidity)

#mean of nat.var. # 2 for columns
apply(climate.nat.var, 2,mean)
#sd of nat.var
apply(climate.nat.var, 2, sd)
#mean of capt.eff
apply(climate.capt.eff, 2, mean, na.rm = TRUE)
#mean of capt.eff
apply(climate.capt.eff, 2, sd, na.rm = TRUE)

climate.all <- nat.var %>%
  filter(day %in% c(15:24)) %>%
  select(headwind, air.temp, humidity)

apply(climate.all, 2, mean)
apply(climate.all, 2, sd)

caught.byday <- caught.hour.day %>%
  group_by(cgt_day) %>%
  summarise(mean.hdw = mean(headwind, na.rm = FALSE),
            mean.art = mean(air.temp, na.rm = FALSE),
            mean.hmd = mean(humidity, na.rm = FALSE),
            sd.hdw = sd(headwind, na.rm = FALSE),
            sd.art = sd(air.temp, na.rm = FALSE),
            sd.hmd = sd(humidity, na.rm = FALSE),
            max.hdw = max(headwind, na.rm = FALSE),
            min.hdw = min(headwind, na.rm = FALSE)
            ) %>%
            na.omit()


caught.byday

day0 <- caught.hour.day %>%
          filter(cgt_day == 0)

day1 <- caught.hour.day %>%
          filter(cgt_day == 1)

t.test(day0$headwind, day1$headwind, paired = FALSE, var.equal = FALSE, conf.level = 0.95)
t.test(day0$air.temp, day1$air.temp, paired = FALSE, var.equal = FALSE, conf.level = 0.95)
t.test(day0$humidity, day1$humidity, paired = FALSE, var.equal = FALSE, conf.level = 0.95)

```

```{r}
t.test(climate.nat.var$headwind, climate.capt.eff$headwind, paired = FALSE, var.equal = FALSE, conf.level = 0.95)
t.test(climate.nat.var$air.temp, climate.capt.eff$air.temp, paired = FALSE, var.equal = FALSE, conf.level = 0.95)
t.test(climate.nat.var$humidity, climate.capt.eff$humidity, paired = FALSE, var.equal = FALSE, conf.level = 0.95)
```

### duration difference 
```{r}
raw.flight.metrics <- nat.var %>%
  group_by(species) %>%
  summarise(mean.v1 = mean(v1, na.rm = FALSE),
            mean.v2 = mean(v2, na.rm = FALSE),
            mean.acc = mean(acc, na.rm = FALSE),
            mean.avg.v = mean(avg.v, na.rm = FALSE),
            mean.t1 = mean(t1, na.rm = FALSE),
            mean.t2 = mean(t2, na.rm = FALSE),
            mean.t3 = mean((t1+t2), na.rm = FALSE),
            sd.v1 = sd(v1, na.rm = FALSE),
            sd.v2 = sd(v2, na.rm = FALSE),
            sd.acc = sd(acc, na.rm = FALSE),
            sd.avg.v = sd(avg.v, na.rm = FALSE),
            sd.t1 = sd(t1, na.rm = FALSE),
            sd.t2 = sd(t2, na.rm = FALSE),
            sd.t3 = sd((t1+t2), na.rm = FALSE)
            ) %>%
            na.omit()

raw.flight.metrics




bluti.raw <- nat.var %>%
  filter(species == "bluti") %>%
  mutate(t3 = t1 + t2)

greti.raw <- nat.var %>%
  filter(species == "greti") %>%
  mutate(t3 = t1 + t2)





t.test(bluti.raw$t1, greti.raw$t1, paired = FALSE, var.equal = FALSE, conf.level = 0.95)

t.test(bluti.raw$t2, greti.raw$t2, paired = FALSE, var.equal = FALSE, conf.level = 0.95)

t.test(bluti.raw$t3, greti.raw$t3, paired = FALSE, var.equal = FALSE, conf.level = 0.95)
```
write.csv(raw.flight.metrics, "raw.flight.metrics.csv",row.names = FALSE )


# II. Graphs {.tabset} 
## A. Structural variation {.tabset}
### 1. individual variation
```{r, echo = FALSE, warning = FALSE}
indv_diff_v1 <- ggplot(data= nat.var, aes(x=Ring, y=v1, fill = species)) +
  geom_boxplot(alpha = 1)+ 
  #stat_boxplot(geom = "errorbar", width = 0.5, color = "gray")+
  #stat_summary(fun.y = mean, geom = "point", color = "red", size = 0.9)+
  #stat_summary(fun.data = mean_sd, geom = "errorbar", width = 0.2, color = "red") +
  theme(legend.position = "top") +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())+
   scale_fill_manual(values = c("bluti" ="#619CFF", "greti" = "#00BA38", "marti" = "#D55E00")) +
  ylab("initial velocity (m/s)") 

indv_diff_v2 <- ggplot(data= nat.var, aes(x=Ring, y=v2, fill = species)) +
  geom_boxplot(alpha = 1)+ 
  #stat_boxplot(geom = "errorbar", width = 0.5, color = "gray")+
  #stat_summary(fun.y = mean, geom = "point", color = "red", size = 0.9)+
  #stat_summary(fun.data = mean_sd, geom = "errorbar", width = 0.2, color = "red") +
  theme(legend.position = "none") +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())+
   scale_fill_manual(values = c("bluti" ="#619CFF", "greti" = "#00BA38", "marti" = "#D55E00")) +
  ylab("end velocity (m/s)") 


indv_diff_avg.v <- ggplot(data= nat.var, aes(x=Ring, y=avg.v, fill = species)) +
  geom_boxplot(alpha =1)+ 
  #stat_boxplot(geom = "errorbar", width = 0.5, color = "gray")+
 # stat_summary(fun.y = mean, geom = "point", color = "red", size = 0.9)+
  #stat_summary(fun.data = mean_sd, geom = "errorbar", width = 0.2, color = "red") +
  theme(legend.position = "none") +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())+
   scale_fill_manual(values = c("bluti" ="#619CFF", "greti" = "#00BA38", "marti" = "#D55E00")) +
  ylab("average velocity (m/s)") 
  

indv_diff_acc <- ggplot(data= nat.var, aes(x=Ring, y=acc, fill = species)) +
  geom_boxplot(alpha = 1)+ 
  theme(legend.position = "none") +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())+ 
  scale_fill_manual(values = c("bluti" ="#619CFF", "greti" = "#00BA38", "marti" = "#D55E00")) +
  ylab("acceleration (m/s^2)") 

grid.arrange(indv_diff_v1, indv_diff_v2, indv_diff_avg.v, indv_diff_acc, nrow = 2)
```

### 2. species variation
```{r, echo = FALSE, warning = FALSE}

spec_diff_v1 <- ggplot(data = nat.var, aes(x=species, y=v1, fill = species)) +
  geom_boxplot(alpha =0.7)+ 
  stat_boxplot(geom = "errorbar", width = 0.5, color = "gray")+
  stat_summary(fun = mean, geom = "point", color = "red", size = 0.9)+
  stat_summary(fun.data = mean_sd, geom = "errorbar", width = 0.2, color = "red") +
  theme(legend.position = "none") +
  theme(axis.title.x=element_blank())+
  scale_fill_manual(values = c("bluti" ="#619CFF", "greti" = "#00BA38", "marti" = "#D55E00")) +
  ylab("initial velocity (m/s)")

spec_diff_v2 <- ggplot(data = nat.var, aes(x=species, y=v2, fill = species)) +
  geom_boxplot(alpha =0.7)+ 
  stat_boxplot(geom = "errorbar", width = 0.5, color = "gray")+
  stat_summary(fun = mean, geom = "point", color = "red", size = 0.9)+
  stat_summary(fun.data = mean_sd, geom = "errorbar", width = 0.2, color = "red") +
  theme(legend.position = "none") +
  theme(axis.title.x=element_blank())+
  scale_fill_manual(values = c("bluti" ="#619CFF", "greti" = "#00BA38", "marti" = "#D55E00")) +
  ylab("end velocity (m/s)") 

spec_diff_avg.v <- ggplot(data = nat.var, aes(x=species, y=avg.v, fill = species)) +
  geom_boxplot(alpha =0.7)+ 
  stat_boxplot(geom = "errorbar", width = 0.5, color = "gray")+
  stat_summary(fun = mean, geom = "point", color = "red", size = 0.9)+
  stat_summary(fun.data = mean_sd, geom = "errorbar", width = 0.2, color = "red") +
  theme(legend.position = "none") +
  theme(axis.title.x=element_blank())+
  scale_fill_manual(values = c("bluti" ="#619CFF", "greti" = "#00BA38", "marti" = "#D55E00")) +
  ylab("average velocity (m/s)")
  
spec_diff_acc <- ggplot(data = nat.var, aes(x=species, y=acc, fill = species)) +
  geom_boxplot(alpha =0.7)+ 
  stat_boxplot(geom = "errorbar", width = 0.5, color = "gray")+
  stat_summary(fun = mean, geom = "point", color = "red", size = 0.9)+
  stat_summary(fun.data = mean_sd, geom = "errorbar", width = 0.2, color = "red") +
  theme(legend.position = "none") +
  theme(axis.title.x=element_blank())+
  scale_fill_manual(values = c("bluti" ="#619CFF", "greti" = "#00BA38", "marti" = "#D55E00")) +
  ylab("acceleration (m/s^2)")

grid.arrange(spec_diff_v1, spec_diff_v2, spec_diff_avg.v, spec_diff_acc, nrow =2)
```

### 3. sex x species variation
```{r, include = FALSE}
#find the species without sex determined within blue tit and great tit
all_dates3.sex.na <- all_dates3 %>%
  filter(species %in% c("bluti", "greti")) %>%
  dcast(Ring ~ Sex) %>%
  filter(Var.2 != 0)

all_dates3.sex.na #was initially 4. see if now its changed to 4 ->3. fixed!

table(all_dates3$Sex) #to see how males and femals are nominated

#4 blue tits found and ATR8387 logged as F

for(i in 1:nrow(all_dates3)){
  
  if(all_dates3$Ring[i] == "ATR8387"){
    
    all_dates3$Sex[i] <- "F"
  }
  else{}
  
}

all_dates3.sex.na
```

```{r, echo = FALSE, warning = FALSE, message = FALSE}
#filter marsh tit and 
all_dates3_sex <- all_dates3 %>%
  filter(species %in% c("bluti", "greti"))%>%
  filter(!Ring %in% all_dates3.sex.na$Ring)

#by sex and then facet_wrap by species
sex_diff_v1 <- ggplot(data=all_dates3_sex, aes(x=Sex, y=v1, fill = factor(species))) +
  geom_boxplot(alpha =0.7)+ 
  stat_boxplot(geom = "errorbar", width = 0.5, color = "gray")+
  stat_summary(fun.y = mean, geom = "point", color = "red", size = 0.9)+
  stat_summary(fun.data = mean_sd, geom = "errorbar", width = 0.2, color = "red") +
  theme(legend.position = "none") +
  theme(axis.title.x=element_blank())+
  scale_fill_manual(values = c("bluti" ="#619CFF", "greti" = "#00BA38"), 
                    name = "species") +
  ylab("initial velocity (m/s)") + facet_wrap(~species, labeller = labeller(species = c("greti" = "great tit", "bluti" = "blue tit")))

sex_diff_v2 <- ggplot(data=all_dates3_sex, aes(x=Sex, y=v2, fill = factor(species))) +
  geom_boxplot(alpha =0.7)+ 
  stat_boxplot(geom = "errorbar", width = 0.5, color = "gray")+
  stat_summary(fun.y = mean, geom = "point", color = "red", size = 0.9)+
  stat_summary(fun.data = mean_sd, geom = "errorbar", width = 0.2, color = "red") +
  theme(legend.position = "none") +
  theme(axis.title.x=element_blank())+
  scale_fill_manual(values = c("bluti" ="#619CFF", "greti" = "#00BA38"), 
                    name = "species") +
  ylab("end velocity (m/s)") + facet_wrap(~species, labeller = labeller(species = c("greti" = "great tit", "bluti" = "blue tit")))

sex_diff_avg.v <- ggplot(data=all_dates3_sex, aes(x=Sex, y=avg.v, fill = factor(species))) +
  geom_boxplot(alpha =0.7)+ 
  stat_boxplot(geom = "errorbar", width = 0.5, color = "gray")+
  stat_summary(fun.y = mean, geom = "point", color = "red", size = 0.9)+
  stat_summary(fun.data = mean_sd, geom = "errorbar", width = 0.2, color = "red") +
  theme(legend.position = "none") +
  theme(axis.title.x=element_blank())+
  scale_fill_manual(values = c("bluti" ="#619CFF", "greti" = "#00BA38"), 
                    name = "species") +
  ylab("average velocity (m/s)") + facet_wrap(~species, labeller = labeller(species = c("greti" = "great tit", "bluti" = "blue tit")))

sex_diff_acc <- ggplot(data=all_dates3_sex, aes(x=Sex, y=acc, fill = factor(species))) +
  geom_boxplot(alpha =0.7)+ 
  stat_boxplot(geom = "errorbar", width = 0.5, color = "gray")+
  stat_summary(fun.y = mean, geom = "point", color = "red", size = 0.9)+
  stat_summary(fun.data = mean_sd, geom = "errorbar", width = 0.2, color = "red") +
  theme(legend.position = "none") +
  theme(axis.title.x=element_blank())+
  scale_fill_manual(values = c("bluti" ="#619CFF", "greti" = "#00BA38"), 
                    name = "species") +
  ylab("acceleration (m/s^2)") + facet_wrap(~species, labeller = labeller(species = c("greti" = "great tit", "bluti" = "blue tit")))


grid.arrange(sex_diff_v1, sex_diff_v2, sex_diff_avg.v, sex_diff_acc, nrow = 2)
```

### 4. age variation
```{r, echo = FALSE, warning = FALSE}

#by age and then facet_wrap by species
age_diff_v1 <- ggplot(data = nat.var, aes(x = factor(age), y=v1, fill = factor(species))) +
  geom_boxplot(alpha =0.7)+ 
  stat_boxplot(geom = "errorbar", width = 0.5, color = "gray")+
  stat_summary(fun.y = mean, geom = "point", color = "red", size = 0.9)+
  stat_summary(fun.data = mean_sd, geom = "errorbar", width = 0.2, color = "red") +
  theme(legend.position = "none") +
  theme(axis.title.x=element_blank())+
  scale_fill_manual(values = c("bluti" ="#619CFF", "greti" = "#00BA38", "marti" = "#D55E00")) +
  scale_x_discrete(labels = c('first-year', 'adult')) +
  ylab("initial velocity (m/s)") + 
  facet_wrap(~species, labeller = labeller(species = c("greti" = "great tit", "bluti" = "blue tit", "marti" = "marsh tit"))) 

age_diff_v2 <- ggplot(data = nat.var, aes(x = factor(age), y = v2, fill = factor(species))) +
  geom_boxplot(alpha =0.7)+ 
  stat_boxplot(geom = "errorbar", width = 0.5, color = "gray")+
  stat_summary(fun.y = mean, geom = "point", color = "red", size = 0.9)+
  stat_summary(fun.data = mean_sd, geom = "errorbar", width = 0.2, color = "red") +
  theme(legend.position = "none") +
  theme(axis.title.x=element_blank())+
  scale_fill_manual(values = c("bluti" ="#619CFF", "greti" = "#00BA38", "marti" = "#D55E00")) +
  scale_x_discrete(labels = c('first-year', 'adult')) +
  ylab("end velocity (m/s)") + 
  facet_wrap(~species, labeller = labeller(species = c("greti" = "great tit", "bluti" = "blue tit", "marti" = "marsh tit"))) 

age_diff_avg.v <- ggplot(data = nat.var, aes(x = factor(age), y = avg.v, fill = factor(species))) +
  geom_boxplot(alpha =0.7)+ 
  stat_boxplot(geom = "errorbar", width = 0.5, color = "gray")+
  stat_summary(fun.y = mean, geom = "point", color = "red", size = 0.9)+
  stat_summary(fun.data = mean_sd, geom = "errorbar", width = 0.2, color = "red") +
  theme(legend.position = "none") +
  theme(axis.title.x=element_blank())+
  scale_fill_manual(values = c("bluti" ="#619CFF", "greti" = "#00BA38", "marti" = "#D55E00")) +
  scale_x_discrete(labels = c('first-year', 'adult')) +
  ylab("average velocity (m/s)") + 
  facet_wrap(~species, labeller = labeller(species = c("greti" = "great tit", "bluti" = "blue tit", "marti" = "marsh tit"))) 

age_diff_acc <- ggplot(data = nat.var, aes(x = factor(age), y = acc, fill = factor(species))) +
  geom_boxplot(alpha =0.7)+ 
  stat_boxplot(geom = "errorbar", width = 0.5, color = "gray")+
  stat_summary(fun.y = mean, geom = "point", color = "red", size = 0.9)+
  stat_summary(fun.data = mean_sd, geom = "errorbar", width = 0.2, color = "red") +
  theme(legend.position = "none") +
  theme(axis.title.x=element_blank())+
  scale_fill_manual(values = c("bluti" ="#619CFF", "greti" = "#00BA38", "marti" = "#D55E00")) +
  scale_x_discrete(labels = c('first-year', 'adult')) +
  ylab("acceleration (m/s)") + 
  facet_wrap(~species, labeller = labeller(species = c("greti" = "great tit", "bluti" = "blue tit", "marti" = "marsh tit"))) 

grid.arrange(age_diff_v1, age_diff_v2, age_diff_avg.v, age_diff_acc, nrow = 2)
#add age 
```


## B. climate factors {.tabset}
### 1. headwind
```{r, echo = FALSE, message = FALSE, warning = FALSE}
headwind_v1 <- ggplot(data=nat.var, aes(x = headwind, y= v1, color = species)) +
  geom_point(size = 1, alpha = 0.5) + geom_smooth(method ="lm") +
  ylab("initial velocity (m/s)") + xlab("headwind (m/s)") +
  theme(legend.position = "none") +
 # theme(axis.title.x=element_blank())+
  theme(legend.title = element_blank()) +
  scale_color_manual(values = c("bluti" ="#619CFF", "greti" = "#00BA38", "marti" = "#D55E00"),
                     breaks = c("greti", "marti", "bluti"),
                     labels = c("Great Tit", "Marsh Tit", "Blue Tit")) 


headwind_v2 <- ggplot(data=nat.var, aes(x=headwind, y=v2, color = species)) +
  geom_point(size = 1, alpha =0.5) + geom_smooth(method ="lm") +
  ylab("end velocity (m/s)") + xlab("headwind (m/s)") +
  theme(legend.position = "none") +
  theme(axis.title.x=element_blank())+
  scale_color_manual(values = c("bluti" ="#619CFF", "greti" = "#00BA38", "marti" = "#D55E00")) 

headwind_avg.v <- ggplot(data=nat.var, aes(x=headwind, y=avg.v, color = species)) +
  geom_point(size = 1, alpha = 0.5) + geom_smooth(method ="lm") +
  ylab("average velocity (m/s)") + xlab("headwind (m/s)") +
  theme(legend.position = "none") +
  theme(axis.title.x=element_blank())+
  scale_color_manual(values = c("bluti" ="#619CFF", "greti" = "#00BA38", "marti" = "#D55E00")) 

headwind_acc <- ggplot(data=nat.var, aes(x=headwind, y=acc, color = species)) +
  geom_point(size = 1, alpha = 0.5) + geom_smooth(method ="lm") +
  ylab("acceleration (m/s^2)") + xlab("headwind (m/s)") +
  theme(legend.position = "none") +
  #theme(axis.title.x=element_blank())+
  scale_color_manual(values = c("bluti" ="#619CFF", "greti" = "#00BA38", "marti" = "#D55E00")) 

grid.arrange(headwind_v1, headwind_acc, nrow = 1)

```

### 2. air temperature
```{r, echo = FALSE, message = FALSE, warning = FALSE}
air.temp_v1 <- ggplot(data=all_dates3.climate, aes(x = air.temp, y=v1, color = species)) +
  geom_point(size = 1, alpha = 0.1) + geom_smooth(method ="lm") +
  ylab("iniital velocity (m/s)") + xlab("air temperature (C)") +
  theme(legend.position = "none") +
   scale_color_manual(values = c("bluti" ="#619CFF", "greti" = "#00BA38", "marti" = "#D55E00"))

air.temp_v2 <- ggplot(data=all_dates3.climate, aes(x = air.temp, y=v2, color = species)) +
  geom_point(size = 1, alpha =0.1) + geom_smooth(method ="lm") +
  ylab("end velocity (m/s)") + xlab("air temperature (C)") +
  theme(legend.position = "none") +
   scale_color_manual(values = c("bluti" ="#619CFF", "greti" = "#00BA38", "marti" = "#D55E00"))

air.temp_avg.v <- ggplot(data=all_dates3.climate, aes(x = air.temp, y=avg.v, color = species)) +
  geom_point(size = 1, alpha = 0.1) + geom_smooth(method ="lm") +
  ylab("average velocity (m/s)") + xlab("air temperature (C)") +
  theme(legend.position = "none") +
   scale_color_manual(values = c("bluti" ="#619CFF", "greti" = "#00BA38", "marti" = "#D55E00"))

air.temp_acc <- ggplot(data=all_dates3.climate, aes(x = air.temp, y=acc, color = species)) +
  geom_point(size = 1, alpha = 0.1) + geom_smooth(method ="lm") +
  ylab("acceleration (m/s^2)") + xlab("air temperature (C)") +
  theme(legend.position = "none") +
   scale_color_manual(values = c("bluti" ="#619CFF", "greti" = "#00BA38", "marti" = "#D55E00"))

grid.arrange(air.temp_v1, air.temp_v2, air.temp_avg.v, air.temp_acc, nrow = 2)
```

### 3. humidity
```{r, echo = FALSE, message = FALSE, warning = FALSE}
humidity_v1 <- ggplot(data=all_dates3.climate, aes(x = humidity, y=v1, color = species)) +
  geom_point(size = 1, alpha = 0.1) + geom_smooth(method ="lm") +
  ylab("iniital velocity (m/s)") + xlab("humidity (%)") +
  theme(legend.position = "none") +
   scale_color_manual(values = c("bluti" ="#619CFF", "greti" = "#00BA38", "marti" = "#D55E00"))

humidity_v2 <- ggplot(data=all_dates3.climate, aes(x = humidity, y=v2, color = species)) +
  geom_point(size = 1, alpha =0.1) + geom_smooth(method ="lm") +
  ylab("end velocity (m/s)") + xlab("humidity (%)") +
  theme(legend.position = "none") +
   scale_color_manual(values = c("bluti" ="#619CFF", "greti" = "#00BA38", "marti" = "#D55E00"))

humidity_avg.v <- ggplot(data=all_dates3.climate, aes(x = humidity, y=avg.v, color = species)) +
  geom_point(size = 1, alpha = 0.1) + geom_smooth(method ="lm") +
  ylab("average velocity (m/s)") + xlab("humidity (%)") +
  theme(legend.position = "none") +
   scale_color_manual(values = c("bluti" ="#619CFF", "greti" = "#00BA38", "marti" = "#D55E00"))

humidity_acc <- ggplot(data=all_dates3.climate, aes(x = humidity, y=acc, color = species)) +
  geom_point(size = 1, alpha = 0.1) + geom_smooth(method ="lm") +
  ylab("acceleration (m/s^2)") + xlab("humidity (%)") +
  theme(legend.position = "none") +
   scale_color_manual(values = c("bluti" ="#619CFF", "greti" = "#00BA38", "marti" = "#D55E00"))

grid.arrange(humidity_v1, humidity_v2, humidity_avg.v, humidity_acc, nrow = 2)
```

## C. diurnal variation 
day effect not included in linear or linear mixed models because there was no difference among days until mist-netting happened.
```{r, echo = FALSE, message = FALSE, warning = FALSE}
#fight x hour of day 
plot_v1.all_dates3.spec <- all_dates3 %>% group_by(species, hour) %>%
  summarise(mean = mean(v1, na.rm=FALSE), 
            sd = sd(v1, na.rm=FALSE),
            upper = mean + sd, 
            lower = mean - sd)


plot_v2.all_dates3.spec <- all_dates3 %>% group_by(species, hour) %>%
  summarise(mean = mean(v2, na.rm=FALSE), 
            sd = sd(v2, na.rm=FALSE),
            upper = mean + sd, 
            lower = mean - sd)

plot_avg.v.all_dates3.spec <- all_dates3 %>% group_by(species, hour) %>%
  summarise(mean = mean(avg.v, na.rm=FALSE), 
            sd = sd(avg.v, na.rm=FALSE),
            upper = mean + sd, 
            lower = mean - sd)

plot_acc.all_dates3.spec <- all_dates3 %>% group_by(species, hour) %>%
  summarise(mean = mean(acc, na.rm=FALSE), 
            sd = sd(acc, na.rm=FALSE),
            upper = mean + sd, 
            lower = mean - sd)

all_dates3_summary <- all_dates3 %>% group_by(species) %>% summarise()

#v1
byspecies_alldates_v1 <-
ggplot() + 
  geom_jitter(data = all_dates3, aes(x=hour, y=v1, colour=species), 
              alpha=0.7)  +
  geom_ribbon(data=plot_v1.all_dates3.spec, 
              aes(x=hour, ymax=upper, ymin=lower, fill=species), 
              alpha = 0.5) + 
  geom_line(data=plot_v1.all_dates3.spec, 
            aes(x=hour, y=mean, colour=species, group=species), 
           size=1.5) +
   scale_color_manual(values = c("bluti" ="#56B4E9", "greti" = "#009E73", "marti" ="#D55E00"))  +
  scale_fill_manual(values = c("bluti" ="#619CFF", "greti" = "#00BA38", "marti" = "#D55E00")) + #shade 
  scale_x_continuous(breaks=c(7:18)) +
  ylab("initial velocity (m/s)") +
  theme(legend.position = "none")


#v2
byspecies_alldates_v2 <-
ggplot() + 
  geom_jitter(data = all_dates3, aes(x=hour, y=v2, colour=species), 
              alpha=0.7)  +
  geom_ribbon(data=plot_v2.all_dates3.spec, 
              aes(x=hour, ymax=upper, ymin=lower, fill=species), 
              alpha = 0.5) + 
  geom_line(data=plot_v2.all_dates3.spec, 
            aes(x=hour, y=mean, colour=species, group=species), 
           size=1.5) +
   scale_color_manual(values = c("bluti" ="#56B4E9", "greti" = "#009E73", "marti" ="#D55E00"))  +
  scale_fill_manual(values = c("bluti" ="#619CFF", "greti" = "#00BA38", "marti" = "#D55E00")) + #shade 
  scale_x_continuous(breaks=c(7:18)) +
  ylab("end velocity (m/s)") +
  theme(legend.position = "none")



#Average velocity 
byspecies_alldates_avg.v <-
ggplot() + 
  geom_jitter(data = all_dates3, aes(x=hour, y=avg.v, colour=species), 
              alpha=0.7)  +
  geom_ribbon(data=plot_avg.v.all_dates3.spec, 
              aes(x=hour, ymax=upper, ymin=lower, fill=species), 
              alpha = 0.5) + 
  geom_line(data=plot_avg.v.all_dates3.spec, 
            aes(x=hour, y=mean, colour=species, group=species), 
           size=1.5) +
   scale_color_manual(values = c("bluti" ="#56B4E9", "greti" = "#009E73", "marti" ="#D55E00"))  +
  scale_fill_manual(values = c("bluti" ="#619CFF", "greti" = "#00BA38", "marti" = "#D55E00")) + #shade 
  scale_x_continuous(breaks=c(7:18)) +
  ylab("average velocity (m/s)") +
  theme(legend.position = "none")

#Average acceleration
byspecies_alldates_acc <-
ggplot() + 
  geom_jitter(data = all_dates3, aes(x=hour, y=acc, colour=species), alpha=0.7) +
  geom_ribbon(data=plot_acc.all_dates3.spec, 
              aes(x=hour, ymax=upper, ymin=lower, fill=species), 
              alpha = 0.5) + 
  geom_line(data=plot_acc.all_dates3.spec, 
            aes(x=hour, y=mean, colour=species, group=species), 
            size=1.5)+
 scale_color_manual(values = c("bluti" ="#56B4E9", "greti" = "#009E73", "marti" ="#D55E00"))  +
  scale_fill_manual(values = c("bluti" ="#619CFF", "greti" = "#00BA38", "marti" = "#D55E00")) + #shade 
 scale_x_continuous(breaks=c(7:18)) +
 ylab("acceleration (m/s^2)") +
  theme(legend.position = "none")

grid.arrange(byspecies_alldates_v1, byspecies_alldates_v2, byspecies_alldates_avg.v, byspecies_alldates_acc, nrow =2)


```

# III. Linear Mixed Model analysis 
We chose linear mixed model to account for individual variation in flight speed. We chose headwind from climate factors by its relationship with flight and correlation with air temperature and humidity. 

## A. Model selection 
### 1. checking correlations
```{r, echo = FALSE, warning = FALSE}
colnames(all_dates3.climate)
cor.chart = chart.Correlation(all_dates3.climate[,34:38], histogram=TRUE, pch=19)

```

### 2. Proposed Mixed Linear Models.
#### with headwind
```{r}
lmer.v1 = lmer(v1 ~ species  + Sex  + age + (-headwind) + (1|Ring), data = nat.var)
lmer.v2 = lmer(v2 ~species  + Sex  + age + (-headwind) + (1|Ring), data = nat.var)
lmer.avg.v = lmer(avg.v ~ species  + Sex  + age + (-headwind) + (1|Ring), data = nat.var)
lmer.acc = lmer(acc ~ species  + Sex  + age + (-headwind) + (1|Ring), data = nat.var)

```

#### with air temperature
```{r}
lmer.v1 = lmer(v1 ~ species  + Sex  + age + air.temp + (1|Ring), data = nat.var)
lmer.v2 = lmer(v2 ~species  + Sex  + age + air.temp + (1|Ring), data = nat.var)
lmer.avg.v = lmer(avg.v ~ species  + Sex  + age + air.temp + (1|Ring), data = nat.var)
lmer.acc = lmer(acc ~ species  + Sex  + age + air.temp + (1|Ring), data = nat.var)

```



### 3. preparing dataset without outliers {.tabset}
We then identified 15 outliers using Rosner's Test(1983) and prepared a dataset excluding outliers to see the sensitivity of the models to the outliers. All the outliers were included in the dataset in the end as we could not confirm that these outliers were from measurement errors. 
```{r, warning = FALSE, message = FALSE, results = FALSE}
test.v1 <- rosnerTest(nat.var$v1, k=3)
test.v1$all.stats

test.v2 <- rosnerTest(nat.var$v2, k=10)
test.v2$all.stats

test.avg.v<- rosnerTest(nat.var$avg.v, k=13)
test.avg.v$all.stats

test.acc <- rosnerTest(nat.var$acc, k=3)
test.acc$all.stats

#collecting outlier row numbers
out.v1 <- test.v1$all.stats %>% 
  filter(Outlier == "TRUE") %>%
  select(Obs.Num)

out.v2 <- test.v2$all.stats %>% 
  filter(Outlier == "TRUE") %>%
  select(Obs.Num)

out.avg.v <- test.avg.v$all.stats %>% 
  filter(Outlier == "TRUE") %>%
  select(Obs.Num)

out.acc <- test.acc$all.stats %>% 
  filter(Outlier == "TRUE") %>%
  select(Obs.Num)
```

#### outliers
```{r}
outliers <- rbind(out.v1, out.v2, out.avg.v, out.acc) %>% distinct()
outliers #15 outliers 
```

#### without outliers
```{r}
nat.var.out <- nat.var[-c(outliers$Obs.Num),]
dim(nat.var.out) #1419 observations
```

## B. Linear mixed model assumptions 
### 1. normality tests {.tabset} 
```{r}
# Computes the default residuals (raw conditional)
raw_cond.v1 <- compute_redres(lmer.v1)
raw_cond.v2 <- compute_redres(lmer.v2)
raw_cond.avg.v <- compute_redres(lmer.avg.v)
raw_cond.acc <- compute_redres(lmer.acc)
```
#### initial velocity
```{r}
shapiro.test(raw_cond.v1) #p <0.05
```
#### end velocity
```{r}
shapiro.test(raw_cond.v2) #p <0.05
```
#### average velocity
```{r}
shapiro.test(raw_cond.avg.v) #p <0.05
```
#### end velocity
```{r}
shapiro.test(raw_cond.acc) #p <0.05
```


### 2. checking assumption graphs {.tabset}
#### initial velocity  
```{r}
plot(nat.var$headwind, nat.var$v1) #linearity
plot(nat.var$air.temp, nat.var$v1) #linearity

hist(raw_cond.v1) #normality in histogram
plot_redres(lmer.v1) #heteroskedasticity
plot_resqq(lmer.v1) #normality of residuals
plot_ranef(lmer.v1) #normality of random effects residuals
```
  
#### end velocity
```{r}
plot(nat.var$headwind, nat.var$v2)
plot(nat.var$air.temp, nat.var$v2) #wider range
hist(raw_cond.v2) #normality in histogram
plot_redres(lmer.v2) #heteroskedasticity
plot_resqq(lmer.v2) #normality of residuals. not great..
plot_ranef(lmer.v2) #normality of random effects residuals
```

#### average velocity
```{r}
plot(nat.var$headwind, nat.var$avg.v)
plot(nat.var$air.temp, nat.var$avg.v)
hist(raw_cond.avg.v) #normality in histogram
plot_redres(lmer.avg.v) #heteroskedasticity
plot_resqq(lmer.avg.v) #normality of residuals. #not great..
plot_ranef(lmer.avg.v) #normality of random effects residuals
```

#### acceleration
```{r}
plot(nat.var$headwind, nat.var$acc)
plot(nat.var$air.temp, nat.var$acc) #better linearity
hist(raw_cond.acc) #normality in histogram
plot_redres(lmer.acc) #heteroskedasticity
plot_resqq(lmer.acc) #normality of residuals. great. 
plot_ranef(lmer.acc) #normality of random effects residuals
```


## C. Model summary {.tabset}
### initial velocity
```{r}
library(lmerTest)
Anova(lmer.v1, type = 3, test.statistic = "F") #Kenward-Rodger df
summary(lmer.v1)
lsmeans(lmer.v1, pairwise~species)
```
### end velocity
```{r}
Anova(lmer.v2, type = 3, test.statistic = "F") #nothing significant
summary(lmer.v2)
```
### average velocity 
```{r}
Anova(lmer.avg.v, type = 3, test.statistic = "F")
summary(lmer.avg.v)
```
### acceleration
```{r}
Anova(lmer.acc, type = 3, test.statistic = "F")
summary(lmer.acc)
lsmeans(lmer.acc, pairwise~ species) 
```

# IV. Best model selection {.tabset}
## A. automatic model averaging 
### with headwind 
#### v1 and acc
##### no interaction
```{r}
set.seed(583)
summary(model.avg(dredge(lmer(v1 ~ species  + Sex + age + headwind + (1|Ring), data = nat.var, na.action = "na.fail")), subset = delta <  5))

set.seed(951)
summary(model.avg(dredge(lmer(acc ~ species  + Sex + age + headwind + (1|Ring), data = nat.var, na.action = "na.fail")), subset = delta <  5))
```
##### interaction
```{r}
set.seed(583)
summary(model.avg(dredge(lmer(v1 ~ species  * Sex * age + headwind + (1|Ring), data = nat.var, na.action = "na.fail")), subset = delta <  5))

set.seed(951)
summary(model.avg(dredge(lmer(acc ~ species  * Sex * age + headwind + (1|Ring), data = nat.var, na.action = "na.fail")), subset = delta <  5))


```
#### v2 and avg.v
- tried log(response variable), but still same result of just headwind.
```{r}
set.seed(86)
summary(model.avg(dredge(lmer(log(v2) ~ species  + Sex + age + headwind + (1|Ring), data = nat.var, na.action = "na.fail")), subset = delta <  10))

set.seed(124)
summary(model.avg(dredge(lmer(log(avg.v) ~ species  + Sex + age + headwind + (1|Ring), data = nat.var, na.action = "na.fail")), subset = delta <  5))

```
- without logging
```{r}
set.seed(86)
summary(model.avg(dredge(lmer(v2 ~ species  + Sex + age + headwind + (1|Ring), data = nat.var, na.action = "na.fail")), subset = delta <  10))

set.seed(124)
summary(model.avg(dredge(lmer(avg.v ~ species  + Sex + age + headwind + (1|Ring), data = nat.var, na.action = "na.fail")), subset = delta <  5))

```

### with air.temp 
- In avg.v and v2 models, (Null) with no predictors better than any other combination of predictors by deltaAIC of 4.20 and 5.07 respectively. 
```{r}
summary(model.avg(dredge(lmer(v1 ~ species  + Sex + age + air.temp + (1|Ring), data = nat.var, na.action = "na.fail")), subset = delta <  10))

summary(model.avg(dredge(lmer(v2 ~ species  + Sex + age + air.temp + (1|Ring), data = nat.var, na.action = "na.fail")), subset = delta <  10)) #null AICc higher by 11 delta AIC

summary(model.avg(dredge(lmer(avg.v ~ species  + Sex + age + air.temp + (1|Ring), data = nat.var, na.action = "na.fail")), subset = delta <  10)) #null AICc higher by 7.62 delta AIC 

summary(model.avg(dredge(lmer(acc ~ species  + Sex + age + air.temp + (1|Ring), data = nat.var, na.action = "na.fail")), subset = delta <  10))

```

## B. final models
```{r}
lmer.v1.fin <- lmer(v1 ~ species + (1|Ring), data = nat.var)
lmer.v2.fin <- lmer(v2 ~ I(-headwind) + (1|Ring), data = nat.var)
lmer.avg.v.fin <- lmer(avg.v ~ I(-headwind) + (1|Ring), data = nat.var)
lmer.acc.fin <- lmer(acc ~ species  + Sex + (1|Ring), data = nat.var)

#without outliers
lmer.v1.fin.out <- lmer(v1 ~ species + (1|Ring), data = nat.var.out)
lmer.v2.fin.out <- lmer(v2 ~ I(-headwind) + (1|Ring), data = nat.var.out)
lmer.avg.v.fin.out <- lmer(avg.v ~ I(-headwind) + (1|Ring), data = nat.var.out)
lmer.acc.fin.out <- lmer(acc ~ species  + Sex + (1|Ring), data = nat.var.out)

```
## C. checking best model assumptions {.tabset}  
Acceleration is skipped as the best model was the full model, acceleration only had one outlier, and the full model assumptions were well validated.   
  
## D. normality tests {.tabset}   
```{r}
# Computes the default residuals (raw conditional)
raw_cond.v1.fin <- compute_redres(lmer.v1.fin)
raw_cond.v1.fin.out <- compute_redres(lmer.v1.fin.out)

raw_cond.v2.fin <- compute_redres(lmer.v2.fin)
raw_cond.v2.fin.out <- compute_redres(lmer.v2.fin.out)

raw_cond.avg.v.fin <- compute_redres(lmer.avg.v.fin)
raw_cond.avg.v.fin.out <- compute_redres(lmer.avg.v.fin.out)

#however checking the one without the outliers
raw_cond.acc.fin <- compute_redres(lmer.acc.fin)
raw_cond.acc.fin.out <- compute_redres(lmer.acc.fin.out)

```
### initial velocity
```{r}
shapiro.test(raw_cond.v1.fin)
shapiro.test(raw_cond.v1.fin.out)
```

### end velocity
```{r}
shapiro.test(raw_cond.v2.fin)
shapiro.test(raw_cond.v2.fin.out)
```
### average velocity
```{r}
shapiro.test(raw_cond.avg.v.fin)
shapiro.test(raw_cond.avg.v.fin.out)
```
### end velocity
```{r}
shapiro.test(raw_cond.acc.fin)
shapiro.test(raw_cond.acc.fin.out)
```

### acceleration
```{r}
shapiro.test(raw_cond.acc.fin)
shapiro.test(raw_cond.acc.fin.out)
```

## E. checking assumptions - no outliers vs. outliers {.tabset}      
### initial velocity      
```{r}
hist(raw_cond.v1.fin) #normality in histogram
hist(raw_cond.v1.fin.out)
plot_redres(lmer.v1.fin) #heteroskedasticity
plot_redres(lmer.v1.fin.out)
plot_resqq(lmer.v1.fin) #normality of residuals
plot_resqq(lmer.v1.fin.out)
plot_ranef(lmer.v1.fin) #normality of random effects residuals
plot_ranef(lmer.v1.fin.out) 
```


### end velocity  
- model without outliers is better but still deviation in normality of residuals graph
```{r}
hist(raw_cond.v2.fin) #normality in histogram
hist(raw_cond.v2.fin.out)
plot_redres(lmer.v2.fin) #heteroskedasticity
plot_redres(lmer.v2.fin.out)
plot_resqq(lmer.v2.fin) #normality of residuals
plot_resqq(lmer.v2.fin.out) 
plot_ranef(lmer.v2.fin) #normality of random effects residuals
plot_ranef(lmer.v2.fin.out) 
```
#### trying data transformation
```{r}

lmer.v2.fin.out.log <- lmer(log(v2) ~ headwind + (1|Ring), data = nat.var.out)

raw_cond.v2.fin.out.log <- compute_redres(lmer.v2.fin.out.log)

hist(raw_cond.v2.fin.out.log)
plot_redres(lmer.v2.fin.out.log)
plot_resqq(lmer.v2.fin.out.log) #still not good
plot_ranef(lmer.v2.fin.out.log) #fine

```

### average velocity   
- model without outliers is better but still deviation in normality of residuals graph
```{r}
hist(raw_cond.avg.v.fin) #normality in histogram
hist(raw_cond.avg.v.fin.out)
plot_redres(lmer.avg.v.fin) #heteroskedasticity
plot_redres(lmer.avg.v.fin.out)
plot_resqq(lmer.avg.v.fin) #normality of residuals
plot_resqq(lmer.avg.v.fin.out)
plot_ranef(lmer.avg.v.fin) #normality of random effects residuals
plot_ranef(lmer.avg.v.fin.out)
```
#### trying data transformation
```{r}

lmer.avg.v.fin.out.log <- lmer(log(avg.v) ~ headwind + (1|Ring), data = nat.var.out)

raw_cond.avg.v.fin.out.log <- compute_redres(lmer.avg.v.fin.out.log)

plot_redres(lmer.avg.v.fin.out.log)
plot_resqq(lmer.avg.v.fin.out.log) #not good
plot_ranef(lmer.avg.v.fin.out.log) #good

```


## F. Best model summary {.tabset}  
### initial velocity 
```{r}
#contrasts = list(Ring = contr.sum, cgt_day = contr.sum))
Anova(lmer.v1.fin, type = 3, test.statistic = "F") #species 
Anova(lmer.v1.fin.out, type = 3, test.statistic = "F")
summary(lmer.v1.fin)
tab_model(lmer.v1.fin)

lsmeans(lmer.v1.fin, pairwise~species)
lsmeans.v1 <- lsmeans(lmer.v1.fin, pairwise~species)
```
### end velocity 
```{r}
Anova(lmer.v2.fin, type = 3, test.statistic = "F") #species and headwind
Anova(lmer.v2.fin.out, type = 3, test.statistic = "F")
summary(lmer.v2.fin)
tab_model(lmer.v2.fin)

```
### average velocity
```{r}
Anova(lmer.avg.v.fin, type = 3, test.statistic = "F") #species and headwind
Anova(lmer.avg.v.fin.out, type = 3, test.statistic = "F")
summary(lmer.avg.v.fin)
tab_model(lmer.avg.v.fin)

```
### acceleration
```{r}
Anova(lmer.acc.fin, type = 3, test.statistic = "F") #species and Sex
Anova(lmer.acc.fin.out, type = 3, test.statistic = "F")
summary(lmer.acc.fin)
tab_model(lmer.acc.fin)

#by species
lsmeans(lmer.acc.fin, pairwise~species)

#by sex
lsmeans(lmer.acc.fin, pairwise~Sex)
lsmeans.acc <- lsmeans(lmer.acc.fin, pairwise~species)
```

## G. plotting species difference
### v1
```{r}
plot.v1.means <- data.frame(lsmeans.v1$lsmeans)

plot.v1.lsmeans <- 
  ggplot() +  
  geom_point(data = plot.v1.means, aes(x = species, y = lsmean, color = species), alpha = 1) +
  geom_errorbar(data = plot.v1.means, aes(x = species, ymin = lower.CL, ymax = upper.CL, color = species, width = 0.5),alpha = 1) +
  scale_color_manual(values = c("bluti" ="black", "greti" = "black", "marti" = "black")) +
  scale_x_discrete(labels = c("Blue Tit","Great Tit","Marsh Tit")) +
  ylab(expression(initial~velocity~(m/s))) +
  ylim(2.25,3.25)+
  theme(axis.title.x=element_blank()) + #getting rid of x-axis title
  theme(legend.position = "none")  #getting rid of legend
```
### acc
```{r}

plot.acc.means <- data.frame(lsmeans.acc$lsmeans)

plot.acc.lsmeans <- 
  ggplot() +  
  geom_point(data = plot.acc.means, aes(x = species, y = lsmean, color = species), alpha = 1) +
  geom_errorbar(data = plot.acc.means, aes(x = species, ymin = lower.CL, ymax = upper.CL, color = species, width = 0.5),alpha = 1) +
  scale_color_manual(values = c("bluti" = "#619CFF", "greti" = "#D55E00", "marti" = "grey")) +
  scale_x_discrete(labels = c("Blue Tit","Great Tit","Marsh Tit")) +
  ylab(expression(`in`-flight~acceleration~(m/s^2))) +
  ylim(-2,6) +
  theme(axis.title.x=element_blank()) + #getting rid of x-axis title
  theme(legend.position = "none") #getting rid of legend

grid.arrange(plot.v1.lsmeans, plot.acc.lsmeans, nrow = 1)
```
### tarsus length
```{r}
species_tarsus <- nat.var %>% group_by(species) %>%
  summarise(
    Mean_Tarsus = mean(Tarsus, na.rm = TRUE),
    SD_Tarsus = sd(Tarsus, na.rm = TRUE)
  )

species_tarsus
```

# V. Repeatability
## v1
### repeatability
```{r}
v1.rpt <- rpt(v1 ~ species + (1|Ring), data = nat.var, grname = "Ring", datatype = "Gaussian", nboot = 10000, npermut = 0)

v1.rpt

plot.v1.rpt <- plot(v1.rpt)

```
#### variance 
```{r}
v1.var <- rpt(v1 ~ species  + Sex + (1|Ring), data = nat.var,grname = "Ring", datatype = "Gaussian", nboot = 10000, npermut = 0,ratio = FALSE)

v1.var

plot.v1.var <- plot(v1.var)

```
## acc
### repeatability
```{r}
acc.rpt <- rpt(acc ~ species  + Sex + (1|Ring), data = nat.var,grname = "Ring", datatype = "Gaussian", nboot = 10000, npermut = 0)

acc.rpt

```
### variance
```{r}
acc.var <- rpt(acc ~ species  + Sex + (1|Ring), data = nat.var,grname = "Ring", datatype = "Gaussian", nboot = 10000, npermut = 0,ratio = FALSE)

acc.var

```
## PLOTS
```{r}
plot(v1.rpt, main = "individual repeatability for v1")
plot(v1.var, main = "individual variance for v1")
plot(acc.rpt, main = "individual repeatability for acc")
plot(acc.var, main = "individual variance for acc")


```

# VI. Response to reviewers

## 1. Raw data graph of individual & species variance 
### v1
```{r}

# Calculate summary statistics: mean and ±1 SD per individual
summary_df <- nat.var %>%
  group_by(Ring, species) %>%
  summarise(
    mean_v1 = mean(v1, na.rm = TRUE),
    sd_v1 = sd(v1, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  mutate(
    sd_lower = mean_v1 - sd_v1,
    sd_upper = mean_v1 + sd_v1
  )

species_summary <- nat.var %>%
  group_by(species) %>%
  summarise(
    mean_v1 = mean(v1, na.rm = TRUE),
    sd_v1 = sd(v1, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  mutate(
    sd_lower = mean_v1 - sd_v1,
    sd_upper = mean_v1 + sd_v1
  )

# Plot
v1_var_graph <- ggplot(nat.var, aes(x = Ring, y = v1, color = species)) +
  geom_jitter(width = 0.2, alpha = 0.5) +  # individual v1 dots
  geom_point(data = summary_df, aes(x=Ring, y = mean_v1, color ="grey"), size = 2, shape = 16, inherit.aes = FALSE) +  # mean points
  geom_errorbar(data = summary_df, aes(x=Ring, ymin = sd_lower, ymax = sd_upper, color = "grey"), width = 0.2, inherit.aes = FALSE) +  # ±1 SD
  
# Species-level overlaying in black
  geom_point(data = species_summary, aes(x = Inf-0.5, y = mean_v1), color = "black", size = 4, shape = 18, inherit.aes = FALSE) +
  geom_errorbar(data = species_summary, aes(x = Inf-0.5, ymin = sd_lower, ymax = sd_upper), color = "black", width = 0.4, inherit.aes = FALSE) +
  facet_wrap(~ species, scales = "free_x", labeller = labeller(
  species = c(
    "bluti" = "Blue Tit",
    "greti" = "Great Tit",
    "marti" = "Marsh Tit" ))) +  # group by species
  theme_minimal() +
  labs(
    x ="",
    y = expression("Initial velocity (m/s)"),
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  theme(
  axis.text.x = element_blank(),     # hides x-axis labels
  axis.ticks.x = element_blank(),    # hides x-axis ticks
  legend.position = "none",
  strip.text = element_text(size = 18),
  axis.title.y = element_text(size = 18)
)

v1_var_graph
```

### acc
```{r}
summary_df <- nat.var %>%
  group_by(Ring, species) %>%
  summarise(
    mean_acc = mean(acc, na.rm = TRUE),
    sd_acc = sd(acc, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  mutate(
    sd_lower = mean_acc - sd_acc,
    sd_upper = mean_acc + sd_acc
  )

species_summary <- nat.var %>%
  group_by(species) %>%
  summarise(
    mean_acc = mean(acc, na.rm = TRUE),
    sd_acc = sd(acc, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  mutate(
    sd_lower = mean_acc - sd_acc,
    sd_upper = mean_acc + sd_acc
  )

# Plot
acc_var_graph <- ggplot(nat.var, aes(x = Ring, y = acc, color = species)) +
  geom_jitter(width = 0.2, alpha = 0.5) +  # individual v1 dots
  geom_point(data = summary_df, aes(x=Ring, y = mean_acc, color ="grey"), size = 2, shape = 16, inherit.aes = FALSE) +  # mean points
  geom_errorbar(data = summary_df, aes(x=Ring, ymin = sd_lower, ymax = sd_upper, color = "grey"), width = 0.2, inherit.aes = FALSE) +  # ±1 SD
  
# Species-level overlaying in black
  geom_point(data = species_summary, aes(x = Inf-0.5, y = mean_acc), color = "black", size = 4, shape = 18, inherit.aes = FALSE) +
  geom_errorbar(data = species_summary, aes(x = Inf-0.5, ymin = sd_lower, ymax = sd_upper), color = "black", width = 0.4, inherit.aes = FALSE) +
  facet_wrap(~ species, scales = "free_x", labeller = labeller(
  species = c(
    "bluti" = "",
    "greti" = "",
    "marti" = "" ))) +  
  theme_minimal() +
  labs(
    x = "",
    y = expression("In-flight acceleration (" * m/s^2 * ")")
  ) +
  #theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(
   # hides x-axis ticks
  axis.text.x = element_blank(),     # hides x-axis labels
  axis.ticks.x = element_blank(),    # hides x-axis ticks
  legend.position = "none",
  axis.title.y = element_text(size = 18)
  #strip.text = element_text(size = 16, face = "bold")
)

acc_var_graph
```

### final plots
```{r}

ggarrange(v1_var_graph, acc_var_graph, nrow =2 )

```

## 2. Non-significance of weather effects
### v1
#### headwind
```{r}
Anova(lmer(v1 ~ species  + Sex  + age + headwind + (1|Ring), data = nat.var), type = 3, test.statistic = "F")

summary(lmer(v1 ~ species  + Sex  + age + headwind + (1|Ring), data = nat.var), type = 3)

min(nat.var$headwind)
max(nat.var$headwind)
mean(nat.var$headwind)
sd(nat.var$headwind)
```

##### acc
```{r}
Anova(lmer(acc ~ species  + Sex  + age + headwind + (1|Ring), data = nat.var), type = 3, test.statistic = "F")

summary(lmer(acc ~ species  + Sex  + age + headwind + (1|Ring), data = nat.var), type = 3)
```
## 3. Are interactions any significant?
### v1
```{r}
Anova(lmer(v1 ~ species*Sex*age + headwind + (1|Ring), data = nat.var), type = 3, test.statistic = "F")

summary(lmer(v1 ~ species*Sex*age + headwind + (1|Ring), data = nat.var), type = 3)
```

### acc
```{r}
Anova(lmer(acc ~ species*Sex*age + headwind + (1|Ring), data = nat.var), type = 3, test.statistic = "F")

summary(lmer(acc ~ species*Sex*age + headwind + (1|Ring), data = nat.var), type = 3)

```
### model averaging with headwind 
#### v1 
```{r}
set.seed(583)
summary(model.avg(dredge(lmer(v1 ~ species  * Sex * age + headwind + (1|Ring), data = nat.var, na.action = "na.fail")), subset = delta <  5))
```
#### acc with superbird
```{r}
set.seed(951)
summary(model.avg(dredge(lmer(acc ~ species  * Sex * age + headwind + (1|Ring), data = nat.var, na.action = "na.fail")), subset = delta <  5)) #(Sex + species + Sex:species)
```
#### v2 and avg.v
- tried log(response variable), but still same result of just headwind.
```{r}
set.seed(86)
summary(model.avg(dredge(lmer(v2 ~ species * Sex * age + headwind + (1|Ring), data = nat.var, na.action = "na.fail")), subset = delta <  10))

set.seed(124)
summary(model.avg(dredge(lmer(avg.v ~ species  * Sex * age + headwind + (1|Ring), data = nat.var, na.action = "na.fail")), subset = delta <  5))

```
### v1 by species, age, and sex. 
```{r}
# filtering out those without sex
nat.var_mf <- nat.var %>%
  filter(Sex != "")

  
summary_df <- nat.var_mf %>%
  group_by(Ring, species, Sex, age) %>%
  summarise(
    mean_v1 = mean(v1, na.rm = TRUE),
    sd_v1 = sd(v1, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  mutate(
    sd_lower = mean_v1 - sd_v1,
    sd_upper = mean_v1 + sd_v1
  )

species_summary <- nat.var_mf %>%
  group_by(species, Sex, age) %>%
  summarise(
    mean_v1 = mean(v1, na.rm = TRUE),
    sd_v1 = sd(v1, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  mutate(
    sd_lower = mean_v1 - sd_v1,
    sd_upper = mean_v1 + sd_v1
  )

plot_v1_agesex <- ggplot(nat.var_mf, aes(x = Ring, y = v1)) +
  geom_jitter(width = 0.2, alpha = 0.5, aes(color = species)) +
  geom_point(data = summary_df, aes(x = Ring, y = mean_v1), color = "grey", size = 2, shape = 16, inherit.aes = FALSE) +
  
  geom_errorbar(data = summary_df, aes(x = Ring, ymin = sd_lower, ymax = sd_upper), color = "grey", width = 0.2,
  inherit.aes = FALSE) +
  
  geom_point(data = species_summary, aes(x = Inf - 0.5, y = mean_v1), color = "black", size = 4, shape = 18,
  inherit.aes = FALSE) +
  
  geom_errorbar(data = species_summary, aes(x = Inf - 0.5, ymin = sd_lower, ymax = sd_upper), color = "black", width = 0.4, inherit.aes = FALSE) +
  facet_grid(species ~ Sex + age, labeller = labeller(
    species = c("bluti" = "Blue Tit", "greti" = "Great Tit", "marti" = "Marsh Tit"),
    Sex = label_value,
    age = c("5" = "first year", "6" = "older")
  )) +
  theme_minimal() +
  labs(
    x = "",
    y = expression("Initial Velocity ("*m/s*")")
  ) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "none"
  )

plot_v1_agesex
```

### acc by species, age, and sex. 
```{r}
# filtering out those without sex
nat.var_mf <- nat.var %>%
  filter(Sex != "")

  
summary_df <- nat.var_mf %>%
  group_by(Ring, species, Sex, age) %>%
  summarise(
    mean_acc = mean(acc, na.rm = TRUE),
    sd_acc = sd(acc, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  mutate(
    sd_lower = mean_acc - sd_acc,
    sd_upper = mean_acc + sd_acc
  )

species_summary <- nat.var_mf %>%
  group_by(species, Sex, age) %>%
  summarise(
    mean_acc = mean(acc, na.rm = TRUE),
    sd_acc = sd(acc, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  mutate(
    sd_lower = mean_acc - sd_acc,
    sd_upper = mean_acc + sd_acc
  )

plot_acc_agesex <- ggplot(nat.var_mf, aes(x = Ring, y = acc)) +
  geom_jitter(width = 0.2, alpha = 0.5, aes(color = species)) +
  geom_point(data = summary_df, aes(x = Ring, y = mean_acc), color = "grey", size = 2, shape = 16, inherit.aes = FALSE) +
  
  geom_errorbar(data = summary_df, aes(x = Ring, ymin = sd_lower, ymax = sd_upper), color = "grey", width = 0.2,
  inherit.aes = FALSE) +
  
  geom_point(data = species_summary, aes(x = Inf - 0.5, y = mean_acc), color = "black", size = 4, shape = 18,
  inherit.aes = FALSE) +
  
  geom_errorbar(data = species_summary, aes(x = Inf - 0.5, ymin = sd_lower, ymax = sd_upper), color = "black", width = 0.4, inherit.aes = FALSE) +
  facet_grid(species ~ Sex + age, labeller = labeller(
    species = c("bluti" = "Blue Tit", "greti" = "Great Tit", "marti" = "Marsh Tit"),
    Sex = label_value,
    age = c("5" = "first year", "6" = "older")
  )) +
  theme_minimal() +
  labs(
    x = "",
    y = expression("In-flight acceleration ("*m/s^2*")")
  ) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "none"
  )

plot_acc_agesex
```

- identify the only great tit adult female
- filter that superbird and see if the acceleration still shows interaction effect with sex and age
```{r}

super_ladybird <- nat.var_mf %>%
  filter((species == "greti") & (Sex == "F") & (age == 6))

super_ladybird$Ring
super_ladybird$acc

nat.var_mf_ordinary <- nat.var_mf %>%
  filter(Ring != "TY68155")

# with the entire dataset
nat.var_superbird <- nat.var %>%
  filter((species == "greti") & (Sex == "F") & (age == 6))


nat.var_normal <- nat.var %>%
  filter(!((species == "greti") & (Sex == "F") & (age == 6)))

dim(nat.var_normal)
```


#### acc model averaging without superbird
```{r}

set.seed(951)
summary(model.avg(dredge(lmer(acc ~ species  * Sex * age + I(-headwind) + (1|Ring), data = nat.var_normal, na.action = "na.fail")), subset = delta <  5)) #(Sex + species + Sex:species)

```
### Model results with and without superbird
```{r}

lmer.acc.fin.super <- lmer(acc ~ species*Sex*age + (1|Ring), data = nat.var)



lmer.acc.fin.normal <- lmer(acc ~ species*Sex*age + (1|Ring), data = nat.var_normal)


```
#### with
##### stats
```{r}
#before
Anova(lmer.acc.fin.super, type = 3, test.statistic = "F")

acc.super <- lsmeans(lmer.acc.fin.super, pairwise~(species:Sex:age))
#summary(lmer.acc.fin.super, type = 3)
```
##### plots
```{r}
acc.super.plot <- as.data.frame(acc.super$lsmeans) %>% 
  tidyr::drop_na() %>%
  dplyr::filter(Sex != "")

acc.super.plot
```

```{r}

# Ensure categorical variables are factors with nice orderings/labels (optional)
acc.super.plot <- acc.super.plot %>%
  mutate(
    species = factor(species, levels = c("bluti", "greti"),
                     labels = c("Blue tit", "Great tit")),
    Sex     = factor(Sex, levels = c("F", "M"),
                     labels = c("Female", "Male")),
    age     = factor(age, levels = c(5, 6), 
                     labels = c("first year", "older"))  # facet as categorical (5, 6)
  )
```
```{r}
# --- Plot: means with 95% CI, dodged by Sex, facetted by age ---
#pd <- position_dodge(width = 0.5)

p <- ggplot(acc.super.plot, aes(x = species, y = lsmean, color = Sex, shape = Sex)) +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL),
                width = 0.15, position = position_dodge(width = 0.5), linewidth = 0.7) +
  geom_point(size = 3, position = position_dodge(width = 0.5)) +
  facet_wrap(~ age, nrow = 1, labeller = labeller(age = label_both)) +
  labs(
    x = "Species",
    y = "Mean (lsmean) with 95% CI",
    color = "Sex",
    shape = "Sex",
    title = "Estimated means and 95% CIs by species, sex, and age"
  ) +
  #scale_color_manual(values = c("Female" = "#3b82f6", "Male" = "#ef4444")) +
  theme_minimal(base_size = 12) +
  theme(
    panel.spacing = unit(1, "lines"),
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    legend.position = "right",
    strip.text = element_text(face = "bold")
  )

p
```
#### without
```{r}
#after
Anova(lmer.acc.fin.normal, type = 3, test.statistic = "F")

#summary(lmer.acc.fin.normal, type = 3)

# the interaction effect disappears when the one extraordinary flight of adult female is excluded.
```
- acc: great tit mean & great tit female mean
```{r}

great_tit_overall <- nat.var_mf_ordinary %>%
  filter(species == "greti") %>%
  summarise(
    mean_acc = mean(acc, na.rm = TRUE),
    sd_acc = sd(acc, na.rm = TRUE)
  )

great_tit_female <- nat.var_mf_ordinary %>%
  filter(species == "greti", Sex == "F") %>%
  summarise(
    mean_acc = mean(acc, na.rm = TRUE),
    sd_acc = sd(acc, na.rm = TRUE)
  )

great_tit_overall
great_tit_female

```

## 4. temporal resolution
- filtering the flights with the same or better temporal resolution of 0.013 s 
```{r}
nat.var$temp_res <- (1/nat.var$v1)*0.03
mean(nat.var$temp_res)
sd(nat.var$temp_res)
min(nat.var$temp_res)
max(nat.var$temp_res)

nat.var_temp_res <- nat.var %>%
  filter(temp_res <= 0.013)
```

### v1
```{r}
Anova(lmer(v1 ~ species  + Sex  + age + headwind + (1|Ring), data = nat.var_temp_res), type = 3, test.statistic = "F")

summary(lmer(v1 ~ species  + Sex  + age + headwind + (1|Ring), data = nat.var_temp_res), type = 3)

```

### acc
```{r}
Anova(lmer(acc ~ species  + Sex  + age + headwind + (1|Ring), data = nat.var_temp_res), type = 3, test.statistic = "F")

summary(lmer(acc ~ species  + Sex  + age + headwind + (1|Ring), data = nat.var_temp_res), type = 3)
```
-species is still significant! 
- however the 0.013s time difference came from the original 1434 flights data. 
```{r}
dim(nat.var_temp_res)
unique(nat.var_temp_res$Ring) #48 individuals 
```
- comparing the exit time again in this dataset
```{r}

new_exit_time <- nat.var_temp_res %>%
  group_by(species) %>%
  summarise(
    mean_t1 = mean(t1, na.rm = TRUE),
    sd_t1 = sd(t1, na.rm = TRUE)
  )
new_exit_time

```


## 5. bodyweight 
- bare in mind these are not bodyweight at the time of flight but rather bodyweight that was ever measured of that individual through mist-netting 
```{r}
# calculate mean and sd of bodyweight per species
bw_species <- nat.var %>%
  group_by(species) %>%
  summarise(
    n = n(),
    mean_bodyweight = mean(Wt, na.rm = TRUE),
    sd_bodyweight = sd(Wt, na.rm = TRUE)
  )
bw_species
```